Refactoring the `nls()` function in R

### 2022-05-29 RefactoringNLS/RefactoringNLS.Rmd
More edits -- now at a review point

---
title: "Refactoring the `nls()` function in R"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "2021-8-20"
output: 
    pdf_document:
        keep_tex: true
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

```{r nlsoutx}
# Croucher example
xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2)
NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
NLSdata<-data.frame(xdata, ydata)
# Try full output version of nls
library(nlspkg) # use the packaged version of nls()
result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE)
# str(result) -- displays large amount of material - suppressed here
#    as it is too wide for the page
result
ls(result) # to list the elements of the output
ls(result$m) # and in particular the "m" object
```

@@

Within `nls()`, consider the 
following FOUR different calling sequences for the same problem, though the
fm2a -- an intuitive choice -- will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing we do
not foresee a resolution for this issue.

```{r log4ways, echo=TRUE}
DNase1 <- subset(DNase, Run == 1)

@@

For `nls()` the concern is demonstrated by a short email correspondence.
Following is an email to John Nash from Doug Bates. This is NOT a criticism
of Prof. Bates work, but a reflection on how difficult it is to develop code in this
subject area and to keep it maintainable. We have experienced similar loss of understanding
for some of our own codes.

```
I'm afraid that I don't know the purpose of the recursive call either.  
I know that I wrote the code to use a closure for the response, covariates, etc., 
but I don't recall anything like a recursive call being necessary.

...

This was in response to an email to Doug bates on Jun 29, 2021 from John Nash.

```
    Thanks.

    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
    <https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R>

    This has the test problem and the expanded code. Around line 367 is where we are
    scratching our heads. The function code (from nlsModel()) is in the commented lines below
    the call. This is

@@

    In both there is the recursive call, which must have a purpose I don't understand.
```

@@

- try to obtain a cleaner structure for the overall `nls()` infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as the new features we would like
  to add.

- try to explain what we do, either in comments or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

@@

Below in "Consistent specifications of partially linear models" we point out that
`nls()` uses a different formula specification from the default for a problem if the 
`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
add `algorithm="plinear"` to the call. Moreover, we believe it makes errors more
likely. We suggest a possible approach to avoiding this issue, but have yet to 
cast this into working code (2021-8-19).

@@

Indexed models clearly have a place in some areas of research. However, the current
approach in `nls()` is awkward. The user must use DIFFERENT formulas depending on 
whether the `plinear` algorithm is chosen. This seems to be related to the need for
that approach to use the `lm()` function or its infrastructure, thus employing 
formulas that omit the parameters of models and simply give the variables or functions
thereof, such as interaction terms.

Users, we believe, are much more likely to be comfortable with fully specified formulas.
And such formulas are needed by nonlinear least squares functions such as `minpack.lm::nlsLM`
and `nlsr::nlxb`. Thus we would like the identification of the linear parameters to be,
if possible, automated. At the very least, we should be able to use a structure like

```
algorithm="plinear(parmx, thetaz)"
```

to allow for the full formula to be used with the linear parameters identified.

## Provide tests

We need suitable tests in order:

@@

### A test runner program

@@

Some considerations for our test scrips:

@@

# Outputs of the project

The project output is available in the Git repository https://github.com/nashjc/RNonlinearLS .

## Formal reports or documentation

- RefactoringNLS.Rmd: this document which will become the main report of the project.

- TestsDoc.Rmd: a survey of testing tools in R. It has more general possibilities 
  and fits into the subject of regression testing, in which case a 
  more extensive literature review will be needed.  Note that this
  document reflects the work in the the "Problem sets and test infrastructure"
  below.

@@

## Informal reports

These are documents used to discuss particular aspects of our work. These are part
of the repository https://github.com/nashjc/RNonlinearLS.

- DerivsNLS.Rmd: a document to explain different ways in which Jacobian information
  is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is
  a DRAFT of a script to provide examples.

- ImproveNLS.bib: a consolidated BibTex bibliography for all documents in this
  project, possibly with wider application to nonlinear least squares in general.

- MachineSummary.Rmd: an investigation of ways to report the characteristics
  and identity of machines running tests. `MachID.R` offers a suggested concise summary function
  to identify a particular computational system used for tests.

- VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and
  the many choices in their implementation for nonlinear least squares. This is still
  a DRAFT at 2021-8-20.

- PkgFromRbase.Rmd: an explanation of the construction of the `nlspkg` from the
  code in R-base.

  <!-- We could think of putting this in R-bloggers or similar. Or if -->
  <!-- we came up with a more general template, then an R-Journal article. -->

- WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on.
  A project diary.

@@

@@

- nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED. 
  The effort showed that the structure of the programs was difficult to follow, 
  undocumented, and unsuited to adding improvements like the Marquardt stabilization. 
  We were able to get a pure-R version of numericDeriv() and rework most of the functions 
  of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project
  formally ends, but collaboration is likely needed with workers who have a deep knowledge
  of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`.

- nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the 
  fix by JN to the relative offset convergence criterion failure when there are small 
  residuals in problems sent to ``nls()`. Now part of R.

- nlsj: A refactoring of the `nls() functionality. `redesign2107.txt` gives some notes.
  This is an interim package for exploration of ideas and will NOT, as it currently
  is presented, become a distributed package.

- nlsralt: this is intended to be a modified version of Nash and Murdoch package 
  `nlsr` with improvements discovered as a result of this project. At the close of
  this project, this is still a work in progress.

@@

A key difference in approach between `nls()` and `nlsr::nlxb()`, possibly the key
difference, is that `nls()` builds a large infrastructure from which the Gauss-Newton
iteration can be executed and other statistical information such as profiles can be
computed, while `nlxb()` returns quite limited information, and in its execution
computes what is needed on an as and when basis. This follows a path that one of us
(JN) established almost 50 years ago with the software that became @cnm79, using
setup, solver, and post-solution analysis phases to computation. Here, that last 
phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr`
package. `nls()` also has some similar functions, but they are much more tied into
the infrastructure that is created mostly in the `nlsModel()` function.

The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable
entangling of the different functions and capabilities. This is to an extent that even after
both the preparation for and carrying out of the present project, we do not feel confident
to explain the code completely, nor to maintain it. However, we have made some progress toward
such a goal. Part of that has been to render the code in R. In this there remain some tasks
to complete, and they are possibly difficult.

@@

To advance the stability and maintainability of R, we believe that the functions that carry out
quite advanced calculations like those achieved by `nls()` should be organized to have their 
interactions as orderly as possible. The aspects of `nls()` that have given us the most trouble
are as follows:

- The functions that compute the residuals and jacobians are frequently computed by presuming
  the current data and parameters for those functions are available in an environment. As long
  as the correct environment is used, this provides a surprisingly short syntax to invoke the
  calculations.

- Weights, subsets, and various contextual controls such as that for the `na.action` that tells
  our code what to do with undefined or missing numbers. Again, this makes for a very simple
  invocation of the calculation, but the context is hidden from the user -- and it can be
  difficult for those of us trying to maintain or improve the code to be certain we have the
  context correct.

- The mixing of R and C code makes for high performance, but without developer documentation 
  leaves programmers with a lot of work to do to adjust the code. We believe in keeping code --
  at least a runnable reference version -- in a single programming language. If necessary, by
  measuring ("profiling") the code, we can find bottlenecks and replace just those slower 
  parts of the reference code.

@@

### Consistent specifications of partially linear models

Ideally, we believe it would be best to specify all models with a complete set of parameters.
That is, the model should be written down for as it would be used, rather than for estimation. 
The `algorithm="plinear"` setting requires that we leave out the linear parameters. This
means that we cannot simply change the `algorithm` argument without also changing the 
`formula`. Moreover, the partially linear parameters appear in the result object
as `.lin1`, `.lin2`, etc.

A possible syntax to provide consistency would be to specify the partially linear parameters
when the `plinear` option is specified. For example, we could use

```
algorithm="plinear(Asym)"
```

which would allow for the partially linear algorithm to function and the output of different
`algorithm` options to be consistent.

Furthermore, we ask if it is possible for us to devise code that will detect linearity, 
rather than expecting the user to provide the special structure and  setting. Work in this
area appears to tbe quite limited, with @Birkisson2013 being seemingly a sole contribution.
Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin 
is about a class of models that have almost nothing in common
with the present issue. The algorithm in @Birkisson2013 will need investigating further,
but we suspect that it is complicated enough that it will not easily be adapted to 
an `nls()` style program. Possibly a simpler algorithm may be found.

>>>>

---
title: "Refactoring the `nls()` function in R"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "2022-5-29"
output: 
    pdf_document:
        keep_tex: true
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

```{r nlsoutx}
# Croucher example
xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2)
NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
NLSdata<-data.frame(xdata, ydata)
# Try full output version of nls
library(nlspkg) # use the packaged version of nls()
result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE)
# We can display the result with several commands
# str(result) -- displays large amount of material - suppressed here
#    as it is too wide for the page
# result # displays the object
# ls(result) # to list the elements of the output
# ls(result$m) # and in particular the "m" object
```

@@

Within `nls()`, consider the 
following FOUR different calling sequences for the same problem, though the
fm2a -- an intuitive choice -- will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing we do
not foresee a resolution for this issue. In the example we have NOT evaluated
the commands to save space.

```{r log4ways, echo=TRUE, eval=FALSE}
DNase1 <- subset(DNase, Run == 1)

@@

For `nls()` the concern is demonstrated by a short email query from John Nash
to Doug Bates and his reply.
This is NOT a criticism of Prof. Bates (or any other) work, but a reflection on 
how difficult it is to develop code in this subject area and to keep it maintainable. 
We have experienced similar loss of understanding for some of our own codes.

```
    ...
    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
    ... has the test problem and the expanded code. Around line 367 is where we are
    scratching our heads. The function code (from nlsModel()) is in the commented 
    lines below the call. This is

@@

    In both there is the recursive call, which must have a purpose I don't understand.
```
The reply was

```
    I'm afraid that I don't know the purpose of the recursive call either.  
    I know that I wrote the code to use a closure for the response, covariates, etc., 
    but I don't recall anything like a recursive call being necessary.
    ...

@@

- to streamline the overall `nls()` infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

- to explain what we do, either in comments or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

@@

We have seen that
`nls()` uses a different formula specification from the default if the 
`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
add `algorithm="plinear"` to the call. This makes errors more
likely.

While not consistent with the `lm()` function and specification of linear models, 
we believe a complete set of parameters should be specified for nonlinear models.
That is, the model formula should be written down as it would be computed, 
rather than for estimation. Thus `y ~ a*x + b` rather than `y ~ x`.

Moreover, we want to avoid the partially linear parameters appearing in the result object
as `.lin1`, `.lin2`, etc., where there actual position in the model is not explicit.

A possible workaround to allow consistency would be to specify the partially linear parameters
when the `plinear` option is specified. For example, we could use `algorithm="plinear(Asym)"`
while keeping the general model formula from the default specification. This would allow 
for the output of different `algorithm` options to be consistent. However, we have not
yet tried to code such a change.

A further complication of model specification in R is that some formulas require
the user to surround a term with `I(  )` to inhibit interpretation of arithmetic 
operators. We would prefer that formulas be usable without this complication, which
has limited documentation and examples.

It is also of interest to consider code that will detect partial linearity, 
rather than expecting the user to provide the special structure and settings. Work in this
area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution
we can find.
Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin 
is about a class of models that have almost nothing in common
with the present issue. The algorithm in @Birkisson2013 will need investigating further,
but we suspect that it is complicated enough that it will not easily be adapted to 
a program like `nls()`. Possibly a simpler algorithm may be found.

@@

Indexed models clearly have a place in some areas of research. 
We need to be able to process models, as the example above, where the
model formula is 
`Length ~ a[Strip] + b[Strip] * exp(-Conc/th)` and `Strip` is the index. Data
`Length` and `Conc` are available for all observations. Parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` depend on the index, which
could be an experimental run, or some other categorization.

Given that `[ ]` are the R method to index arrays, their presence in a model
formula signals indexing. The question is how to perform the correct calculations.

While tedious (??maybe dummies or fastdummies packages make it less so?), 
we could set up the model as a sum of similar models with dummy
variables to turn "on" or "off" the particular case, transforming the model
parameters to simple, non-indexed ones (as does `nls()`).

Wrapper functions to prepare the estimation calculations and present the
results in a form that matches the modelling formula would avoid adding
complexity to `nls()` and other tools.

## Provide tests and use-case examples

We need suitable tests and use-case examples in order:

@@

### A test runner program ?? NLSCompare

@@

Some considerations for our test scripts:

@@

# Progress so far

Much of our work is reported in https://github.com/nashjc/RNonlinearLS .

## Reports and articles

- The present article is the central report of the Google Summer or Code 2021 project
  "Improvements to nls()".

- VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and
  the many choices in their implementation for nonlinear least squares. This is still
  a DRAFT at 2021-8-20. ?? fix

- TestsDoc.Rmd (??citation) is a survey of testing tools in R. It has more 
  general possibilities 
  and fits into the subject of regression testing, in which case a 
  more extensive literature review will be needed.  Note that this
  document reflects the work in the the "Problem sets and test infrastructure"
  below.

@@

- DerivsNLS.Rmd ??citation: a document to explain different ways in which Jacobian information
  is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is
  a DRAFT of a script to provide examples.

- The BibTex bibliography ImproveNLS.bib (??citation or reference) was set up 
  for use with all documents in this project, but has wider application to nonlinear 
  least squares projects in general.

- MachineSummary.Rmd (??citation -- blog post??) is an investigation of ways to report 
  the characteristics
  and identity of machines running tests. `MachID.R` offers a suggested concise summary function
  to identify a particular computational system used for tests.

- @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
  code in R-base.

- As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
  a review of optimization in R. Ideas from the present work have been instrumental
  in the creation of @NashWires22.

<!-- ?? probably don't need to include -->
<!-- - WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on. -->
<!--   A project diary. -->

@@

?? need to clean up

@@

- nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED. 
  The effort showed that the structure of the programs was difficult to follow, 
  undocumented, and unsuited to adding improvements like the Marquardt stabilization. 
  We were able to get a pure-R version of `numericDeriv()` and rework most of the functions 
  of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project
  formally ends, but collaboration is likely needed with workers who have a deep knowledge
  of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`.

<!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  -->
<!--   fix by JN to the relative offset convergence criterion failure when there are small  -->
<!--   residuals in problems sent to ``nls()`. Now part of R. -->

- the safeguarded relative offset convergence criterion that allows small-residual problems
  to ba handled with `nls()` is now in the R base code.

- `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features??
  This is an interim package for exploration of ideas and will NOT, as it currently
  is presented, become a distributed package.

- nlsralt: this is intended to be a modified version of Nash and Murdoch package 
  `nlsr` to examine possible improvements discovered as a result of this project. 
  This is EXPERIMENTAL.

@@

The key difference in approach between `nls()` and `nlsr::nlxb()` is that `nls()` 
builds a large infrastructure from which the Gauss-Newton
iteration can be executed and other statistical information such as profiles can be
computed, while `nlxb()` returns quite limited information, and in its execution
computes what is needed on an "as and when" basis. This follows a path that one of us
(JN) established almost 50 years ago with the software that became @cnm79, using
setup, solver, and post-solution analysis phases to computation. Here, that last 
phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr`
package. `nls()` also has some similar functions, but they are much more tied into
the infrastructure that is created, mostly in the `nlsModel()` function.

The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable
entangling of the different functions and capabilities. Even after
both the preparation for and carrying out of the present project, we do not feel confident
to explain the code completely, nor to maintain it. However, we have made some 
forays into render parts of the code in R, and to understand where difficulties
lie.

@@

To advance the stability and maintainability of R, we believe the program objects (R functions)
that are created by tools such as `nls()` should have minimal interactions and side-effects.
The aspects of `nls()` that have given us the most trouble are as follows:

- The functions that compute the residuals and jacobians are frequently computed by presuming
  the current data and parameters for those functions are available in an environment. As long
  as the correct environment is used, this provides a surprisingly short syntax to invoke the
  calculations. The danger is that the wrong data is accessed if the internal search finds a
  valid name that is not the object we want.

- Weights, subsets, and various contextual controls such as that for the `na.action` that tell
  our code what to do with undefined or missing numbers. Again, this makes for a very simple
  invocation of the calculation, but the context is hidden from the user. It can be
  difficult for those of us trying to maintain or improve the code to be certain we have the
  context correct.

- The mixing of R and C code made for high performance in the past, but without 
  developer documentation, programmers face a lot of work to fix code. We believe in keeping code --
  at least a runnable reference version -- in a single programming language. If necessary, by
  measuring ("profiling") the code, we can find bottlenecks and replace just those slower 
  parts of the reference code.

@@
